Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-05-02 PAPER.tex
edited language to reflect using a subset of the data only; also careful about using the word 'revised PS' since this paper does not revise the A15 PS

The major challenge that faces all 21\,cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{santos_et_al2005, ali_et_al2008, deOliveiraCosta_et_al2008, jelic_et_al2008, bernardi_et_al2009, bernardi_et_al2010, ghosh_et_al2011, pober_et_al2013b, bernardi_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three topics (signal loss, error estimation, and bias) that are essential to investigate 
for a robust 21\,cm power spectrum analysis. We also highlight four power spectrum techniques (fringe-rate filtering, weighting, bootstrapping, jackknife testing) and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER, motivating a revised PAPER-64 power spectrum 
from the lessons learned. A companion paper, Kolopanis et al. (\textit{submitted}), builds off of the methods in this paper to present PAPER-64 results for multiple redshifts and baseline types.

@@

We now apply these lessons to data from the PAPER experiment to make a new analysis of the PAPER-64 dataset originally presented in 
\citetalias{ali_et_al2015} and obtain a revised power spectrum estimate.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0cm 3cm 0cm 3cm},width=\columnwidth]{plots/ant_layout_aspect.pdf}
	\caption{The PAPER-64 antenna layout. We use only the $30$ m East/West baselines for the revised analysis in this 
paper (i.e. the shortest horizontal spacings).}
	\label{fig:ant_layout}
\end{figure}

@@

For our revised analysis, we use $8.1$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (\citetalias{ali_et_al2015} uses a slightly longer RA 
range of $0$-$8.6$ hours; we found that some early LSTs were more severely foreground contaminated). All power spectrum results are produced for a center frequency of 151\,MHz using a width of 10\,MHz ($20$ channels), identical to the analysis in \citetalias{ali_et_al2015}. In the case study in this paper, we only use one baseline type instead of the three as in 
\citetalias{ali_et_al2015}, but Kolopanis et al. (\textit{submitted}) uses all baselines presented in \citetalias{ali_et_al2015}.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_modeloop_2panel.pdf}
	\caption{Power spectra $2\sigma$ upper limits for $k=0.393$\,$h$ Mpc$^{-1}$ for fringe-rate filtered PAPER-64 data. Top: Values 
are shown before (dashed red) and after (solid red) signal loss estimation via our signal injection framework as a function of number of eigenmodes of $\widehat{\textbf{C}}$ that 
are down-weighted. This regularization knob is tuned from $0$ modes on the left (i.e. unweighted) to $21$ modes on the right (i.e. the full inverse 
covariance estimator). Over $\sim3$ orders of magnitude of signal loss results when using empirically estimated inverse covariance weighting. Bottom: Power spectrum upper limits before (dashed red) and after (solid red) signal loss estimation as a function of identity added to the empirical covariance. This regularization knob is tuned from $\gamma = 10^{-4}$ on the right (i.e. very little regularization) to $\gamma = 1$ on the left (see main text for the definition of $\gamma$). Also 
plotted in both panels for comparison are $2\sigma$ power spectrum upper limits for the uniform-weighted case (blue) and inverse variance 
weighted case (black); both are after signal loss estimation. Finally, a theoretical prediction for noise ($2\sigma$ error) is plotted 
as green. In the revised PAPER-64 analysis in this paper, we choose to use a regularization scheme of $\widehat{\textbf{C}}_{\rm eff} \equiv 0.035 \, $Tr($\widehat{\textbf{C}})\textbf{I} + \widehat{\textbf{C}}$ ($\gamma = 0.035$) as a simple example of regularization that minimizes loss, and note that the power spectrum limits using this type of regularization are roughly constant across a large range of values of $\gamma$.}
	\label{fig:sigloss_modeloop}
\end{figure*}

@@

A revised PAPER-64 power spectrum for our subset of data (using only one baseline separation type and $\widehat{\textbf{C}}_{\rm eff}$) is shown in Figure 
\ref{fig:ps1_data}. Again, the solid red curve represents our upper limit on the EoR signal using the full signal injection framework. The uniform weighted case is shown as the black and gray points, which correspond to positive and negative power spectrum values respectively (with 
$2\sigma$ errors bars). It is also shown as an upper limit using the signal injection framework (solid blue). Finally, the gray shaded regions combine the measured uniform weighted power spectrum values and the thermal noise. We show this power spectrum result as one example of how a simple regularization can minimize signal loss, though we also note that this weighting does not produce more stringent limits than the uniform weighted case. In Kolopanis et al. (\textit{submitted}), multiple redshifts are explored and found to benefit from weighting differently.

@@

Re-analysis of the PAPER-64 data included a detailed study using several independently generated noise simulations. What we 
found was that these simulations all agreed but were discrepant with the previous analytic sensitivity calculations. The analytic 
calculation is only an approximation; however, the differences were large enough (factors of $10$ in some cases) to warrant a 
careful investigation. The analytic calculation attempts to combine a large number of pieces of information in an approximate 
way, and when re-considering some of the approximations, we have found there to be large effects. What follows here is an 
accounting of the differences which have been discovered. Our revised theoretical error estimation, which is often plotted as the solid green curve in many of the previous power spectrum plots, is computed with these changes accounted for.

@@

N_{\rm gps}}{2}}$\Big). A revised \citetalias{ali_et_al2015} analysis should therefore use $N_{\rm bls} \sim 32$ instead of $51$, and this change is taken into account in Figure \ref{fig:sense_check}. However, the analysis in this paper and in Kolopanis et al. (\textit{submitted}) does not average baselines into groups ($N_{\rm gps} = 1$), and hence $N_{\rm bls} = 51$ for the revised PAPER-64 analysis.

@@

In addition to mitigating foreground bias at high-$k$'s, two other sources of bias that we actively suppress in the PAPER-64 
analysis are noise bias associated with the squaring of thermal noise and noise bias from crosstalk. In order to avoid the 
former, we filter out certain cross-multiplications when forming $\widehat{q}$ in Equation \eqref{eq:qhat}. Namely, the PAPER-64 
dataset is divided into two halves: even Julian dates and odd Julian dates. Our data vectors are then $\textbf{x}_{even, 1}$ for 
the ``even" dataset and baseline group $1$, $\textbf{x}_{odd, 1}$ for the ``odd" dataset and baseline group $1$, etc. We only form 
$\widehat{q}$ when the two copies of $\textbf{x}$ come from different groups and baselines, never multiplying ``even" with ``even", for 
example, in order to prevent the squaring of the same thermal noise. We recognize that this method results in a hit to our sensitivity. Sensitivity can be gained, for example, by only dividing up the dataset once (either along time or baselines), or by creating more groups (more cross-multiplications), but in this paper we only attempt to revise the \citetalias{ali_et_al2015} analysis (which uses the same groupings).

>>>>

The major challenge that faces all 21\,cm experiments is isolating a small signal that is buried underneath foregrounds and 
instrumental systematics that are, when combined, four to five orders of magnitude brighter \citep[e.g.,][]{santos_et_al2005, ali_et_al2008, deOliveiraCosta_et_al2008, jelic_et_al2008, bernardi_et_al2009, bernardi_et_al2010, ghosh_et_al2011, pober_et_al2013b, bernardi_et_al2013, dillon_et_al2014, kohn_et_al2016}. A clean measurement therefore requires an intimate understanding of the instrument and a rigorous study of data analysis choices. With continual progress being made 
in the field and HERA on the horizon, it is becoming increasingly important to understand how the methods we choose interact 
with each other to affect power spectrum results. More specifically, it is imperative to develop techniques and tests that ensure 
the accuracy and reliability of a potential EoR detection. In this paper, we discuss three topics (signal loss, error estimation, and bias) that are essential to investigate 
for a robust 21\,cm power spectrum analysis. We also highlight four power spectrum techniques (fringe-rate filtering, weighting, bootstrapping, jackknife testing) and their trade-offs, potential 
pitfalls, and connections to the themes. We first approach the themes from a broad perspective, and then perform a detailed 
case study using data from the 64-element configuration of PAPER. In this study we use a subset of the PAPER-64 data to illustrate our revised analysis methods, while a companion paper, Kolopanis et al. (\textit{submitted}), builds off of the methods in this paper to present revised PAPER-64 results for multiple redshifts and baseline types.

@@

We now apply these lessons to data from the PAPER experiment, using a subset of the PAPER-64 dataset from
\citetalias{ali_et_al2015}) in order to illustrate our revised analysis pipeline.

@@

\begin{figure}
	\centering
	\includegraphics[trim={0cm 3cm 0cm 3cm},width=\columnwidth]{plots/ant_layout_aspect.pdf}
	\caption{The PAPER-64 antenna layout. We use only $10$ of the $30$ m East/West baselines for the analysis in this 
paper (i.e. a subset of the shortest horizontal spacings).}
	\label{fig:ant_layout}
\end{figure}

@@

For the analysis in this paper, we use $8.1$ hours of LST, namely an RA range of $0.5$-$8.6$ hours (\citetalias{ali_et_al2015} uses a slightly longer RA 
range of $0$-$8.6$ hours; we found that some early LSTs were more severely foreground contaminated). We also use only $10$ baselines, a subset of the $51$ total East/West baselines used in \citetalias{ali_et_al2015}. All power spectrum results are produced for a center frequency of 151\,MHz using a width of 10\,MHz ($20$ channels), identical to the analysis in \citetalias{ali_et_al2015}. In the case study in this paper, we only use one baseline type instead of the three as in 
\citetalias{ali_et_al2015}, but Kolopanis et al. (\textit{submitted}) uses all baselines presented in \citetalias{ali_et_al2015}.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_modeloop_2panel.pdf}
	\caption{Power spectra $2\sigma$ upper limits for $k=0.393$\,$h$ Mpc$^{-1}$ for fringe-rate filtered PAPER-64 data. Top: Values 
are shown before (dashed red) and after (solid red) signal loss estimation via our signal injection framework as a function of number of eigenmodes of $\widehat{\textbf{C}}$ that 
are down-weighted. This regularization knob is tuned from $0$ modes on the left (i.e. unweighted) to $21$ modes on the right (i.e. the full inverse 
covariance estimator). Over $\sim3$ orders of magnitude of signal loss results when using empirically estimated inverse covariance weighting. Bottom: Power spectrum upper limits before (dashed red) and after (solid red) signal loss estimation as a function of identity added to the empirical covariance. This regularization knob is tuned from $\gamma = 10^{-4}$ on the right (i.e. very little regularization) to $\gamma = 1$ on the left (see main text for the definition of $\gamma$). Also 
plotted in both panels for comparison are $2\sigma$ power spectrum upper limits for the uniform-weighted case (blue) and inverse variance 
weighted case (black); both are after signal loss estimation. Finally, a theoretical prediction for noise ($2\sigma$ error) is plotted 
as green. In the PAPER-64 analysis in this paper, we choose to use a regularization scheme of $\widehat{\textbf{C}}_{\rm eff} \equiv 0.035 \, $Tr($\widehat{\textbf{C}})\textbf{I} + \widehat{\textbf{C}}$ ($\gamma = 0.035$) as a simple example of regularization that minimizes loss, and note that the power spectrum limits using this type of regularization are roughly constant across a large range of values of $\gamma$.}
	\label{fig:sigloss_modeloop}
\end{figure*}

@@

The power spectrum result for our subset of PAPER-64 data (using only one baseline separation type and $\widehat{\textbf{C}}_{\rm eff}$) is shown in Figure 
\ref{fig:ps1_data}. Again, the solid red curve represents our upper limit on the EoR signal using the full signal injection framework. The uniform weighted case is shown as the black and gray points, which correspond to positive and negative power spectrum values respectively (with 
$2\sigma$ errors bars). It is also shown as an upper limit using the signal injection framework (solid blue). Finally, the gray shaded regions combine the measured uniform weighted power spectrum values and the thermal noise. We show this power spectrum result as one example of how a simple regularization can minimize signal loss, though we also note that this weighting does not produce more stringent limits than the uniform weighted case. In Kolopanis et al. (\textit{submitted}), multiple redshifts are explored and found to benefit from weighting differently.

@@

Re-analysis of the PAPER-64 data included a detailed study using several independently generated noise simulations. What we 
found was that these simulations all agreed but were discrepant with the previous analytic sensitivity calculations. The analytic 
calculation is only an approximation; however, the differences were large enough (factors of $10$ in some cases) to warrant a 
careful investigation. The analytic calculation attempts to combine a large number of pieces of information in an approximate 
way, and when re-considering some of the approximations, we have found there to be large effects. What follows here is an 
accounting of the differences which have been discovered. Our revised theoretical error estimation, which is plotted as the solid green curve in many of the previous power spectrum plots, is computed with these changes accounted for.

@@

N_{\rm gps}}{2}}$\Big). A revised \citetalias{ali_et_al2015} analysis should therefore use $N_{\rm bls} \sim 32$ instead of $51$, and this change is taken into account in Figure \ref{fig:sense_check}. However, the analysis in this paper and in Kolopanis et al. (\textit{submitted}) does not average baselines into groups ($N_{\rm gps} = 1$). For the subset of data presented in this paper, $N_{\rm bls} = 10$.

@@

In addition to mitigating foreground bias at high-$k$'s, two other sources of bias that we actively suppress in the PAPER-64 
analysis are noise bias associated with the squaring of thermal noise and noise bias from crosstalk. In order to avoid the 
former, we filter out certain cross-multiplications when forming $\widehat{q}$ in Equation \eqref{eq:qhat}. Namely, the PAPER-64 
dataset is divided into two halves: even Julian dates and odd Julian dates. Our data vectors are then $\textbf{x}_{even, 1}$ for 
the ``even" dataset and baseline $1$, $\textbf{x}_{odd, 1}$ for the ``odd" dataset and baseline $1$, etc. We only form 
$\widehat{q}$ when the two copies of $\textbf{x}$ come from different groups and baselines, never multiplying ``even" with ``even", for 
example, in order to prevent the squaring of the same thermal noise.
